Staged Abstract Interpreters

### 2019-03-23 paper/20_prelim.tex
reorganize paper structure

>>>>

\newcommand{\TLang}{$L_\lambda$}

\section{Preliminaries}

In this section, we first describe the abstract syntax of the language we will
analyze, then present the generic interpreter shared among the four different
semantics, after which, we instantiate the interpreter to the concrete one. Then we
will proceed in two directions: by staging it we obtain a staged concrete interpreter
(Section~\ref{stagedinterp}); by abstracting it we obtain an unstaged abstract
interpreter (Section~\ref{unstaged_abs}).

We choose to use Scala and monad to demonstrate the idea, but the approach
is not restricted to our choice. One can use imperative style and other MSP languages 
(such as MetaOCaml \cite{DBLP:conf/gpce/CalcagnoTHL03, DBLP:conf/flops/Kiselyov14} 
or Template Haskell \cite{Sheard:2002:TMH:636517.636528} \todo{cite the recent one})
to construct such staged abstract interpreters.

### 2019-03-23 paper/40_absinterp.tex
reorganize paper structure

Based on the generic interface we presented in Section~\ref{bg_lang}, 
we now describe the instantiation of an abstract interpreter.
To keep the presentation simple, we use simple abstract domains, and just establish a 
context-/path-/flow-insensitive, store-widened analysis in this section --- indeed, it is coarse
but simple enough to setup a foundation for the staged abstract interpreter we will present later.
In Section~\ref{cfa}, we will see how to regain context-/path-/flow-sensitivity 
for pushdown control-flow analysis; in Section~\ref{cases_imp}, we will see how to instantiate 
an interval abstract domain for numerical analysis.

\paragraph{Abstract Instantiation.}
Our instantiation roughly follows the Abstracting Abstract Machines methodology 
\cite{DBLP:conf/icfp/HornM10, DBLP:journals/jfp/HornM12} that transforms the concrete semantics 
to abstract semantics. The store now maps addresses to sets of abstract values, meaning that an address points
to all possible values that may occur at runtime, where the abstract value is either a closure or 
a single abstract number \texttt{NumV}, which stands for the top element of the number lattice. 
The address space is also constrained to be finite, in the monovariant setting, we simply use variable names 
for address.

@@

\subsection{Stage Polymorphic Lattices} \label{stagedpoly_lat}
To effectively reuse the code between unstaged and staged variants, we also make the lattice structure
stage polymorphic. A lattice type class parameterizes over an element type \texttt{E} and defines
five operations for \texttt{R[E]}, such as meet, join, and ordering relation; again, \texttt{R} 
is the higher-kinded type indicating the binding-time.
\begin{lstlisting}
trait Lattice[E, R[+_]] {
  val bot: R[E];  val top: R[E]
  def $\sqsubseteq$(l1: R[E], l2: R[E]): R[Boolean]
  def $\sqcup$(l1: R[E], l2: R[E]): R[E]
  def $\sqcap$(l1: R[E], l2: R[E]): R[E]
}
\end{lstlisting}

@@

\subsection{Abstract Semantics}
The operations @get@ and @put@ on abstract stores are slightly changed according to
the abstract semantics: @get@ uses the bottom element in the @Value@ lattice if the 
queried address does not exist; @put@ performs the update by joining with the existing values.
function @alloc@ simply invokes the @Addr@ constructor.
\texttt{close} and \texttt{num} lift syntactic literals to our abstract domain, i.e., 
a singleton set that contains only one \texttt{CloV} or \texttt{NumV} object.

@@

For branching, @branch0@ joins the answers from two branches since we are doing a path-insensitive analysis. 
For arithmetics, @prim_eval@ returns the top abstract number element.
It is worth noting that @apply_closure@ is where we handle the non-determinism of applications.
The non-determinism happens because the abstract store maps addresses to sets of abstract values;
when dereferencing an address from the store, we need to explore all possible values to achieve a sound 
analysis. In the case of @apply_closure@, the first argument @fs@ may contains multiple target closures,
so we use a @for@ comprehension over all closures, apply them respectively and form a set result values @vs@;
meanwhile, every time when evaluating the body expression @e@ we use the latest store $\sigma$@_*@
which is updated iteratively during the loop. Thus this is a store-widened analysis.

\subsection{Fixpoint Iteration}
We described the abstract semantics modularly in the last section, however, the abstract 
interpreter may not terminate for some input program. In this section, we use a memoization
technique to ensure its termination. This technique is also widely used under the name of co-inductive caching algorithm
\cite{DBLP:journals/pacmpl/DaraisLNH17, Wei:2018:RAA:3243631.3236800} or truncated depth-first evaluation 
\cite{Rosendahl:AbsIntPL}.

>>>>

Based on the generic interface we presented in Section~\ref{bg_lang}, we now
describe the instantiation of an abstract interpreter. To keep the presentation
simple, we use simple abstract domains, and just establish a
context-/path-/flow-insensitive, store-widened analysis in this section ---
indeed, it is coarse but simple enough to setup a foundation for the staged
abstract interpreter we will present later. In Section~\ref{cfa}, we will see
how to regain context-/path-/flow-sensitivity \todo{path/flow} for pushdown
control-flow analysis; in Section~\ref{cases_imp}, we will see how to
instantiate an interval abstract domain for numerical analysis.

\paragraph{Abstract Instantiation}

Our instantiation roughly follows the Abstracting Abstract Machines methodology
\cite{DBLP:conf/icfp/HornM10, DBLP:journals/jfp/HornM12} that transforms the
concrete semantics to abstract semantics. The store now maps addresses to sets
of abstract values, meaning that an address points to all possible values that
may occur at runtime, where the abstract value is either a closure or a single
abstract number \texttt{NumV}, which stands for the top element of the number
lattice. The address space is also constrained to be finite, in the monovariant
setting, we simply use variable names for address.

@@

\subsection{Stage Polymorphic Lattices} \label{stagedpoly_lat}

To effectively reuse the code between unstaged and staged variants, we also make
the lattice structure stage polymorphic. A lattice type class parameterizes over
an element type \texttt{E} and defines five operations for \texttt{R[E]}, such
as meet, join, and ordering relation; again, \texttt{R} is the higher-kinded
type indicating the binding-time.

\begin{lstlisting}
trait Lattice[E, R[+_]] {
  val bot: R[E];  val top: R[E]
  def $\sqsubseteq$(l1: R[E], l2: R[E]): R[Boolean]
  def $\sqcup$(l1: R[E], l2: R[E]): R[E]
  def $\sqcap$(l1: R[E], l2: R[E]): R[E]
}
\end{lstlisting}

@@

\subsection{Abstract Semantics}

The operations @get@ and @put@ on abstract stores are slightly changed according
to the abstract semantics: @get@ uses the bottom element in the @Value@ lattice
if the queried address does not exist; @put@ performs the update by joining with
the existing values. function @alloc@ simply invokes the @Addr@ constructor.
\texttt{close} and \texttt{num} lift syntactic literals to our abstract domain,
i.e., a singleton set that contains only one \texttt{CloV} or \texttt{NumV}
object.

@@

For branching, @branch0@ joins the answers from two branches since we are doing
a path-insensitive analysis. For arithmetics, @prim_eval@ returns the top
abstract number element. It is worth noting that @apply_closure@ is where we
handle the non-determinism of applications. The non-determinism happens because
the abstract store maps addresses to sets of abstract values; when dereferencing
an address from the store, we need to explore all possible values to achieve a
sound analysis. In the case of @apply_closure@, the first argument @fs@ may
contain multiple target closures, so we use a @for@ comprehension over all
closures, apply them respectively and form a set result values @vs@; meanwhile,
every time when evaluating the body expression @e@ we use the latest store
$\sigma$@_*@ which is updated iteratively during the loop. Thus this is a
store-widened analysis.

\subsection{Fixpoint Iteration}

We described the abstract semantics modularly in the last section, however, the
abstract interpreter may not terminate for some input program. In this section,
we use a memoization technique to ensure its termination. This technique is also
widely used under the name of co-inductive caching algorithm
\cite{DBLP:journals/pacmpl/DaraisLNH17, Wei:2018:RAA:3243631.3236800} or
truncated depth-first evaluation \cite{Rosendahl:AbsIntPL}.

### 2019-03-23 paper/70_eval.tex
reorganize paper structure

\begin{itemize}
    \item \textbf{fib}: Calculates the $n$-th fibonacci number.
    \item \textbf{rsa}: The RSA public key encryption algorithm.
    \item \textbf{kcfa3}: A difficult benchmark for $k$-CFA.
    \item \textbf{church}: Church numerals with additions and mutiplications.
    \item \textbf{fermat}: Fermat and Solovay-Strassen primality testing.
    \item \textbf{kcfa-worst-case-}\textit{n}: 
      Benchmark programs that are supposed to be tough cases for $k$-CFA; 
      the number $n$ indicates the depth of nesting lambda terms.
\end{itemize}

@@

\begin{figure}[h]
\begin{tabular}{@{}llll@{}}
\toprule
    Program            & unstaged   & staged     & $\Delta$ of median \\ \midrule
    fib                & 1.020      & 2.046      & -50.14\%          \\
    rsa                & 35.928     & 145.306    & -75.27\%          \\
    kcfa3              & 2.934      & 5.671      & -48.27\%          \\
    church             & 884.800    & 545.736    & +62.13\%          \\
    fermat             & 15.673     & 68.790     & -77.22\%          \\
    kcfa-worst-16      & 320.689    & 301.564    & +6.34\%           \\
    kcfa-worst-32      & 3,685.615  & 1,854.461  & +98.74\%          \\
    kcfa-worst-64      & 49,422.666 & 14,849.725 & +232.82\%         \\
    \bottomrule
\end{tabular}
\caption{Evaluation result.} \label{evaluation_result}
\end{figure}

>>>>

\begin{itemize}
  %\item \textbf{fib}: Calculates the $n$-th fibonacci number.
  \item \textbf{rsa}: The RSA public key encryption algorithm.
  \item \textbf{kcfa3}: A difficult benchmark for $k$-CFA.
  \item \textbf{church}: Church numerals with additions and mutiplications.
  \item \textbf{fermat}: Fermat and Solovay-Strassen primality testing.
  \item \textbf{kcfa-worst-case-}\textit{n}: 
    Benchmark programs that are supposed to be tough cases for $k$-CFA; 
    the number $n$ indicates the depth of nesting lambda terms.
\end{itemize}

@@

\begin{figure}[h]
\begin{tabular}{@{}llll@{}}
\toprule
    Program            & unstaged   & staged     & $\Delta$ of median \\ \midrule
  % fib                & 1.020      & 2.046      & -50.14\%          \\
    rsa                & 35.928     & 145.306    & -75.27\%          \\
    kcfa3              & 2.934      & 5.671      & -48.27\%          \\
    church             & 884.800    & 545.736    & +62.13\%          \\
    fermat             & 15.673     & 68.790     & -77.22\%          \\
    kcfa-worst-16      & 320.689    & 301.564    & +6.34\%           \\
    kcfa-worst-32      & 3,685.615  & 1,854.461  & +98.74\%          \\
    kcfa-worst-64      & 49,422.666 & 14,849.725 & +232.82\%         \\
    \bottomrule
\end{tabular}
\caption{Evaluation result.} \label{evaluation_result}
\end{figure}
